Generating Empirically Optimized Numerical Software from MATLAB Prototypes

### 2008-12-11 doc/papers/iccs09/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@384 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\begin{figure}[htbp]
\centering
\includegraphics[width=0.7\textwidth]{figures/process.eps}
\caption{Overview of the code generation and tuning process, optional steps are indicated with dashed arrows.}
\label{fig:process}
\end{figure}
To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.  To begin, we have developed a compiler that converts a MATLAB script into simple C code.
%~\cite{Siek}. 
After generating the C code from the high-level MATLAB prototype, we (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} %developed at Ohio State University
to optimize for coarse-grained parallelism and locality simultaneously. Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by our extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters. Orio then performs an empirical search to select the best among the multiple optimized code variants.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.7\textwidth]{figures/compile.eps}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.7\textwidth]{figures/gemver-dataflow.eps}

@@

\begin{figure}[hbtp]
  \centering
  \includegraphics[width=\textwidth]{figures/refine-mv-dot.eps}

@@

\begin{figure}[tb]
\vspace{-.1in}
\begin{center}
\includegraphics[width=.65\textwidth]{figures/orio.eps}
\end{center}
\vspace{-.25in}
\caption{Overview of the Orio empirical tuning process.}
\label{fig:orio}
\end{figure}

@@

We evaluated our approach by running experiments on an Intel Xeon
workstation and the Blue Gene/P at Argonne. The Intel machine has dual
quad-core E5462 Xeon processors (8 cores total) running at 2.8 GHz
(1600 MHz FSB) with 2 GB RAM, running Ubuntu 8.04. Intel C compiler
(v10.1) was used with \texttt{-O3} option (and
\texttt{-parallel}/\texttt{-openmp} for automatic/manual
parallelization). Each node of the Blue Gene/P has four 850 MHz
PowerPC 450 processors with a dual floating-point unit and 2 GB total
memory per node, running a proprietary operating system. On Blue
Gene/P, we used IBM XLC compiler (v9.0), with \texttt{-O3 -qstrict
-qarch=450d -qtune=450 -qhot} options (and
\texttt{-qsmp=auto}/\texttt{-qsmp=noauto} for automatic/manual
parallelization).

@@

\begin{table}[htb] 
\vspace{-.1in} 
\caption{Composed BLAS operations used in our experiments}
\label{tbl:blas-ops}
\small
\centering 
\vspace{-.1in} 
\begin{tabular}{|c|c|c|c|} 
\hline 
Name & Input & Output & C code from MATLAB compiler \\
\hline 
\hline 
VADD & $\overrightarrow{w}$,$\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$ &
$\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ 
\hline 
ATAX & $A$,$\overrightarrow{x}$ & $\overrightarrow{y}$ & 
$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ 
\hline 
BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
& & & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\
\hline 
 & ${\bf A}$,$a$,$b$, & ${\bf B}$, & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
& $\overrightarrow{y}$,$\overrightarrow{z}$ & & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\
\hline 
\end{tabular} 
\vspace{-.2in} 
\end{table}

@@

\begin{minipage}[b]{.515\textwidth}
\includegraphics[width=\textwidth]{figures/vadd_bgp.eps}
\end{minipage}
&
\begin{minipage}[b]{.515\textwidth}
\includegraphics[width=\textwidth]{figures/atax.eps}
\end{minipage}\\

@@

\begin{minipage}[b]{.515\textwidth}
\includegraphics[width=\textwidth]{figures/gemver.eps}
\end{minipage}
&
\begin{minipage}[b]{.515\textwidth}
\includegraphics[width=\textwidth]{figures/bicgkernel.eps}
\end{minipage}\\

>>>>

\begin{figure}[htbp]
\centering
\includegraphics[width=0.7\textwidth]{figures/process.png}
\caption{Overview of the code generation and tuning process, optional steps are indicated with dashed arrows.}
\label{fig:process}
\end{figure}
To generate highly optimized code from a MATLAB prototype of the composed BLAS operation, we follow a three-step approach, illustrated in Figure~\ref{fig:process}.  To begin, we have developed a compiler that converts a MATLAB script into simple C code.
%~\cite{Siek}. 
After generating the C code from the high-level MATLAB prototype, we (optionally) use the source-to-source automatic parallelization tool PLuTo\cite{Pluto} %developed at Ohio State University
to optimize for coarse-grained parallelism and locality simultaneously. Using the results of the PLuTo analysis, we insert annotations into the C code, which are then processed by our extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters. Orio then performs an empirical search to select the best among the multiple optimized code variants.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.7\textwidth]{figures/compile.png}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.7\textwidth]{figures/gemver-dataflow.png}

@@

\begin{figure}[hbtp]
  \centering
  \includegraphics[width=\textwidth]{figures/refine-mv-dot.png}

@@

\begin{figure}[tb]
\vspace{-.1in}
\begin{center}
\includegraphics[width=.65\textwidth]{figures/orio.png}
\end{center}
\vspace{-.25in}
\caption{Overview of the Orio empirical tuning process.}
\label{fig:orio}
\end{figure}

@@

We evaluated our approach by running experiments on an Intel Xeon
workstation and the Blue Gene/P at Argonne. The Intel machine has dual
quad-core E5462 Xeon processors (8 cores total) running at 2.8 GHz
(1600 MHz FSB) with 2 GB RAM, running Ubuntu 8.04. Intel C compiler
(v10.1) was used with \texttt{-O3} option (and
\texttt{-parallel}/\texttt{-openmp} for automatic/manual
parallelization, respectively). Each node of the Blue Gene/P has four
850 MHz PowerPC 450 processors with a dual floating-point unit and 2
GB total memory per node, running a proprietary operating system. On
Blue Gene/P, we used IBM XLC compiler (v9.0), with \texttt{-O3
-qstrict -qarch=450d -qtune=450 -qhot} options (and
\texttt{-qsmp=auto}/\texttt{-qsmp=noauto} for automatic/manual
parallelization, respectively).

@@

\begin{table}[htb] 
\vspace{-.1in} 
\caption{Composed BLAS operations used in our experiments}
\label{tbl:blas-ops}
\small
\centering 
\vspace{-.1in} 
\begin{tabular}{|c|c|c|c|} 
\hline 
Name & Input & Output & C code from MATLAB compiler \\
\hline 
\hline 
VADD & $\overrightarrow{w}$,$\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$ &
$\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ 
\hline 
ATAX & $A$,$\overrightarrow{x}$ & $\overrightarrow{y}$ & 
$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ 
\hline 
 & ${\bf A}$,$a$,$b$, & ${\bf B}$, & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
& $\overrightarrow{y}$,$\overrightarrow{z}$ & & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\
\hline 
BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
& & & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\
\hline 
\end{tabular} 
\vspace{-.2in} 
\end{table}

@@

\begin{minipage}[b]{.515\textwidth}
\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
\end{minipage}
&
\begin{minipage}[b]{.515\textwidth}
\includegraphics[width=\textwidth]{figures/atax.png}
\end{minipage}\\

@@

\begin{minipage}[b]{.515\textwidth}
\includegraphics[width=\textwidth]{figures/gemver.png}
\end{minipage}
&
\begin{minipage}[b]{.515\textwidth}
\includegraphics[width=\textwidth]{figures/bicgkernel.png}
\end{minipage}\\
